### 2022-02-08 symposium.tex
Type and Null Safety for Regexes

\noindent
Here, our implementation instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which it then uses in |unapply|'s return type to obtain a precise representation of the regex's capture groups.

@@

The main challenge...

>>>>

\noindent
Here, we instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capture groups.

@@

The purpose of our type-level component is two-fold:

\begin{enumerate}
  \item Identify the capture groups of the input regular expression.
  \item Determine which of those capture group are optional, that is, whether or not the regex engine can assign null values for each of those groups.
\end{enumerate}

\noindent
The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| element, depending of the optionality analysis.

We present our implementation incrementally, starting from simple but incomplete solution, and progressively building up towards our final solution.
The first version of our type-level program is limited to capture group identification.
The implementation is straightforward:

\begin{lstlisting}
import compiletime.ops.string.{Length, CharAt}
import compiletime.ops.int.+

type Compile[R <: String] =
  Reverse[Loop[R, 0, Length[R], EmptyTuple]]

type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
  Lo match
    case Hi => Acc
    case _   => CharAt[R, Lo] match
      case '('  => Loop[R, Lo + 1, Hi, String *: Acc]
      case '\\' => Loop[R, Lo + 2, Hi, Acc]
      case _ => Loop[R, Lo + 1, Hi, Acc]
\end{lstlisting}

\noindent
The |Length| and |CharAt| types are special cased by the compiler: when their first argument is a literal singleton string, the compiler evaluates those types via their corresponding term-level implementation.
Similarly, the |+| type from |compiletime.ops.int| allows us to manipulate integer just like we would at the term level.
The |EmptyTuple| and |*:| are Scala~3's new generic representation of tuples.
Those types remove the need for heterogeneous lists and allow us to manipulate tuples directly.

Attentive readers might have noticed that our handling of backslash, regex's escape character, might lead to off-by-one errors.
Be reassured, this is not an oversight!
Regular expressions with trailing backslashes result in runtime exceptions (reported as "unexpected internal error").
Even though detecting there type of error at compile time is out of the scope of our implementation, doing so is still desirable.
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into an call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error.

Our first implementation of |Compile| treats every opening parethesis as the start of a capture group, which does no do honor syntax of Java's regular expression.
Indeed, Java also support several other special constructs which start with an opening parethesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, and must all be discriminated from named and unnamed-capturing groups, |(?<name>X)| \& |(X)|.

\regexIsCapturing
% type IsCapturing[R <: String, At <: Int] =
%   CharAt[R, At] match
%     case "?" => CharAt[R, At + 1] match
%       case "<" => CharAt[R, At + 2] match
%         case "!" => false // zero-width negative lookbehind
%         case _ => true    // named-capturing group
%       case _ => false     // other non-capturing special constructs
%     case _ => true        // unnamed-capturing group
